---
title: "Cleaning file"
output: html_document
date: "2022-12-09"
---

````{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
````

## Import Libraries

```{r}
library(dplyr)
library(tidyverse)
library(lubridate)
```

## Import Datasets from data folder

```{r}
#READ DATA
flu <- read.csv("data/raw/flu_vax_coverage.csv")
education <- read.csv("data/raw/education.csv")
pop <- read.csv("data/raw/population.csv")
income <- read.csv("data/raw/income.csv")
```

## Data Transformation:

demo_df will contain transformed/engineered demographic data
```{r}
demo_df <- data.frame()
```


### Transform flu vaccine data by state:

Notice that date data is separated into 2 columns: Season.Survey.Year contains year info (the number after year is not month!), and Month contains month info. 

Added columns: ym: date of the observation, in yyyy-mm-01 format (day doesn't matter)

```{r}
flu$ym <- as.Date(paste(stringi::stri_sub(flu$`Season.Survey.Year`, to = 4),"-", flu$Month, "-1", sep = ""))
flu$`Estimate....` <- as.numeric(flu$`Estimate....`) #NA's!
```

### Transform education attainment data by state:

Added columns: high_school_higher_pop -counts total number of people who attained a degree > hight school over 18

Columns appended to demo_df: high_school_higher_pop
               
```{r}
rowname <- education$Label..Grouping. #store the names for later use
education <- mutate_all(education[str_detect(colnames(education),"Total..Estimate")],
                        function(x){as.numeric(str_replace_all(x,",",""))})%>% t() %>% as.data.frame() #Convert all entries to numeric
rownames(education)<- as.list(lapply(str_split(rownames(education),"\\.\\."),
                                     function(x){str_replace_all(x[1],"\\."," ")})) #Transform state names
colnames(education)<- rowname


#add a column: high_school_higher_pop which counts total number of people who attained a degree > high  school over 18
hs <- as.data.frame(rowSums(education[c(4:6,15)]))
colnames(hs)<- "high_school_higher_pop"
education$high_school_higher_pop<-  hs$high_school_higher_pop

#store the added column in demo_df
demo_df=education
demo_df=demo_df[69]
```

### Transform pop data by state:

* Added columns: 
  + children_pop - number of people <= 14; 
  + children_ratio - ratio of children with respect to total pop; 
  + elder_ratio - ratio of people >= 65

* Columns added to demo_df:
  + pop_over_18
  + high_school_education_ratio (=pop_over_18/tot pop over 18)
  + children_pop
  + children_ratio
  + elder_pop
  + elder_ratio 
  + tot_pop: total population of that state


```{r}
rowname <- pop$Label..Grouping.
pop <- mutate_all(pop[str_detect(colnames(pop),"Total..Estimate")],
                        function(x){as.numeric(str_replace_all(x,",",""))})%>% t() %>% as.data.frame() #Convert all entries to numeric
rownames(pop)<- as.list(lapply(str_split(rownames(pop),"\\.\\."),
                                     function(x){str_replace_all(x[1],"\\."," ")})) #Transform state names
colnames(pop)<- rowname
demo_df$pop_over_18<- pop$`        18 years and over`
```

Calculate the ratio:

```{r}
demo_df<-demo_df %>% mutate(high_school_education_ratio=high_school_higher_pop/pop_over_18)
```

We also consider the ratio of elderly people (>= 65 years old) and children (<= 14 years old) in the population:

```{r}
pop<- pop %>% mutate(children_pop = pop$`        Under 5 years`+pop$`        5 to 9 years` +pop$`        10 to 14 years`,
                     children_ratio = children_pop/`Total population`,
                     elder_ratio = pop$`        65 years and over`/`Total population`)
demo_df$tot_pop<- pop$`Total population`
demo_df$children_pop <- pop$children_pop
demo_df$children_ratio<- pop$children_ratio
demo_df$elder_pop <- pop$`        65 years and over`
demo_df$elder_ratio <- pop$elder_ratio
```

### Transform household income data by state:

Note: US data is missing in this dataset

* Added columns:
  + household_poverty_rate: ratio of households who earn less than 14,999 USD/Year
  + household_wealth_rate: ratio of households who earn more than 150,000 USD/Year 
* Columns added to demo_df:
  + household_poverty_rate
  + household_wealth_rate

```{r}
rowname <- income$Label..Grouping.
income <- income[str_detect(colnames(income),"Households..Estimate")] %>% t() %>% as.data.frame()
rownames(income)<- as.list(lapply(str_split(rownames(income),"\\.\\."),
                                     function(x){str_replace_all(x[1],"\\."," ")})) #Transform state names
colnames(income)<- rowname

#Convert some entries to numeric
income$`Median income (dollars)` <- as.numeric(str_replace_all(income$`Median income (dollars)`,",","")) 
income$`Mean income (dollars)`<-as.numeric(str_replace_all(income$`Mean income (dollars)`,",",""))
demo_df$`median_household_income` <- append(income$`Median income (dollars)`,NA)
demo_df$`mean_household_income` <- append(income$`Mean income (dollars)`,NA)
```

Next, we add household poverty / wealth rates in each state. We define poverty line to be 14,999 USD/Year, and "walth line" to be 150,000 USD/Year.

```{r}
income <- income  %>% mutate(household_poverty_rate= as.numeric(str_replace_all(income$`    Less than $10,000`,"%",""))/100+
                                          as.numeric(str_replace_all(income$`    $10,000 to $14,999`,"%",""))/100,
                             household_wealth_rate=as.numeric(str_replace_all(income$`    $150,000 to $199,999`,"%",""))/100+
                                          as.numeric(str_replace_all(income$`    $200,000 or more`,"%",""))/100)
demo_df$household_poverty_rate = append(income$household_poverty_rate,NA) #append NA to US data
demo_df$household_wealth_rate = append(income$household_wealth_rate,NA)
```

### Final changes to demo_df: merge with flu coverage data

Finally, we append flu shot coverage data by state to demo_df:

```{r}
#FLU SHOT COVERAGE BY STATE IN 2021
shot_by_state = flu 

#50 states + Washington DC
shot_by_state <- shot_by_state %>% filter(Vaccine=="Seasonal Influenza", #H1N1 vaccines are only provided in 2009
                     flu$`Geography.Type`=="States/Local Areas", 
                     Dimension.Type =="Age",
                     ym == as.Date("2021-05-01"), #Missing June! Only 11 Months!
                     Dimension=="≥6 Months") %>% 
  group_by(Geography)%>%
  arrange(ym) %>%
  summarise("n_months"=n(),
            "Mean Vaccine Coverage in 2021 (Percentage)" = mean(Estimate...., na.rm=TRUE))%>%
  ungroup()

demo_df$flu_shot_coverage_2021 <- append(shot_by_state$`Mean Vaccine Coverage in 2021 (Percentage)`,c(NA,NA))
```

### Output clean demo_df:

We have finished building demo_df.The final result:
Since we are missing some values for Puerto Rico, you can consider only include the first 51 rows.
```{r}
demo_df<-head(demo_df,51)
demo_df <- tibble::rownames_to_column(demo_df, "state")

write.csv(demo_df, "data/clean/demo_df.csv", row.names = FALSE)
```



### US flu coverage Data by year(FOR TIME SERIES)

aggregate all records that is focusing on the whole US.

```{r}

us_by_year = flu

us_by_year <- us_by_year%>% filter(Vaccine=="Seasonal Influenza", #H1N1 vaccines are only provided in 2009
                     flu$`Geography.Type`=="HHS Regions/National", 
                     Geography=="United States",
                     Dimension.Type =="Age",
                     Dimension=="≥6 Months") %>% arrange(ym)

write.csv(us_by_year, "data/clean/us_by_year.csv", row.names=FALSE)

```


```{r}

covid2 <- read.csv("data/raw/United_States_COVID-19_Cases.csv")

covid2 <- covid2[-which(covid2$tot_cases == 0 |is.na(covid2$tot_cases)),]
covid2 <- covid2 %>% dplyr::select(submission_date, state, new_case)
covid_time2 <- group_by(covid2,submission_date)
new_cases <- summarize(covid_time2, sum(new_case))
new_cases <- rename(new_cases, new_case='sum(new_case)')
new_cases$submission_date <- as.Date(new_cases$submission_date, format = "%m/%d/%Y")
new_cases <- arrange(new_cases,submission_date)
ggplot(new_cases, aes(submission_date, new_case)) + 
  geom_smooth(method = "loess", span = .05, se = FALSE)


```
```{r}

coverage_rate <- rename(us_by_year, estimation='Estimate....')
coverage_rate$ym <- as.Date(coverage_rate$ym, format = "%m/%d/%Y")
ggplot(coverage_rate, aes(ym, estimation)) + 
  geom_smooth(method = "loess", span = .05, se = FALSE)


```


```{r}
#state based flu data

test <-
  flu %>%
    filter(Vaccine=="Seasonal Influenza") %>%
    filter(Geography.Type == "States/Local Areas",
    Dimension.Type =="Age",
    Dimension=="≥6 Months") %>% arrange(ym)

write.csv(test,"data/clean/coverage_by_state.csv")


```


```{r}

international <- read.csv("data/raw/international_rates.csv")



```

